### 2024-05-04 paper.tex
initial commit

%%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
%%\documentclass[sigplan,nonacm]{acmart}
\documentclass[sigplan, nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\graphicspath{{pictures/}}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
	\providecommand\BibTeX{{%
			\normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
%%\setcopyright{acmcopyright}
\setcopyright{none}
%%\acmJournal{PACMPL}
%%\acmVolume{1}
%%\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
%%\acmArticle{1}
%%\acmYear{2021}
%%\acmMonth{9}
%%\acmDOI{}
%%\copyrightyear{2018}
%%\acmYear{2018}
%%\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
%%\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium%% on Neural
%%  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
%%\acmPrice{15.00}
%%\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}
\usepackage{xspace}
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.78]{beramono}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{stmaryrd}
%\usepackage{unicode-math}
%\usepackage{MnSymbol}
\usepackage{wasysym}
\usepackage{color}
\usepackage{xcolor,colortbl}
\usepackage{url}
\usepackage{listings}
\usepackage{paralist}
%\usepackage[compact]{titlesec}
\usepackage[font={small}]{caption}
\usepackage{wrapfig}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{flushend}
\usepackage{bcprules}
\usepackage{textcomp}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,calc,arrows.meta,arrows,decorations}
\usepackage{pdfpages}
\usepackage{cleveref}
\usetikzlibrary{matrix}
\usepackage{xspace}
\definecolor{light-gray}{gray}{0.85}
\usepackage{stackengine}
\usepackage{mathpartir}
\input{macros}

%%
%% end of the preamble, start of the body of the document source.
\newcommand{\tool}{\textsc{GenSym}\xspace}
\newcommand{\klee}{\textsc{KLEE}\xspace}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\lb}{\{~}
\newcommand{\rb}{~\}}

\newcommand{\Typ}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\Ast}[1]{\ensuremath{\textsf{\text{#1}}}}
\newcommand{\Def}[1]{\ensuremath{\mathrm{#1}}}
\newcommand{\mit}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\msf}[1]{\ensuremath{\mathsf{#1}}}

\newcommand{\lang}{\textsf{SimpLLVM}\xspace}

\newcommand{\Sem}[3][]{\ensuremath{\llbracket {#2} \rrbracket_{#3}^{#1}}}
\newcommand{\SSem}[2]{\ensuremath{\llbracket {#1} \rrbracket_{#2}^{\#}}}
\newcommand{\State}{\mathbb{S}}
\newcommand{\Address}{\mathcal{A}}
\newcommand{\Mem}{\mathbb{M}}
\newcommand{\Value}{\mathcal{V}}
\newcommand{\Loc}{\msf{Loc}}
\begin{document}
\sloppy

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Efficient code generation for loop-free IR}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ruiqi Gao}
\email{gao606@purdue.edu}
\affiliation{%
  \institution{Purdue University}
  \city{Lafayette}
  \state{Indiana}
  \country{USA}
  \postcode{47901}
}

\begin{abstract}
\cite{abeysinghe2024rhyme, abeysingherhyme}
\end{abstract}

\keywords{Loop scheduling, Loop fusion, Code generation}

\maketitle

\section{Introduction}
\subsection{Paper Structure}
The paper is organized as follows: \par \par
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput

### 2024-05-05 paper.tex
finish Section 3

@@
 %% end of the preamble, start of the body of the document source.
-\newcommand{\tool}{\textsc{GenSym}\xspace}
-\newcommand{\klee}{\textsc{KLEE}\xspace}
+\newcommand{\rhyme}{\text{Rhyme}\xspace}
 \newcommand{\ul}[1]{\underline{#1}}
@@
 \begin{abstract}
-\cite{abeysinghe2024rhyme, abeysingherhyme}
+  \rhyme is a declarative multi-paradigm query language designed for querying nested data structures like JSON and performing tensor computation.
+\rhyme performs query optimization and code generation by generating a loop-free, branch-free intermediate representation.
+\rhyme IR does not explicitly capture program structures; instead, all program structures are implicitly captured by dependencies.
+This IR design allows for flexible code scheduling and optimization.
+However, it requires non-trivial reasoning and scheduling of the compiler to generate correct and performant code.\par
+
+  In this paper, we present an efficient code generation algorithm for \rhyme IR.
+Our algorithm performs aggressive loop scheduling and loop fusion to generate optimized code while ensuring program correctness.
+It conducts multiple dependency analyses to ensure correctness and uses a heuristic-guided greedy algorithm to perform aggressive loop fusion.
 \end{abstract}
@@
 \section{Introduction}
+Rhyme\cite{abeysinghe2024rhyme, abeysingherhyme}, is a declarative language targeting the querying of nested data structures such as JSON and tensors.
+Rhyme draws inspiration from existing query languages like JQ \cite{jq}, GraphQL \cite{graphql}, and XQuery \cite{xquery} for querying nested data structures.
+Similar to Datalog \cite{datalog} and JSONPath \cite{jsonpath}, Rhyme employs logic variables * to explicitly represent iterators over data sources.
+Inspired by functional languages like Verse \cite{verse} and Curry \cite{curry}, Rhyme uses unification to represent aggregation.
+Additionally, Rhyme can express tensor computations mathematically, akin to Einstein notation \cite{einops, einsumblog, tensor_comprehensions}.
+
+Rhyme supports the expression of both traditional data processing operations such as aggregation, joins, and group-by, and mathematical computations like matrix multiplication and transpose in a unified language.
+Rhyme also includes internal support for data visualization, allowing users to perform various workloads in a single language without relying on multiple backends.
+
+To perform query optimization and code generation, Rhyme generates an intermediate representation (IR).
+Traditional IRs like LLVM \cite{lattner2004llvm} statically represent the program structure within the IR itself, with all statements scoped within blocks.
+Therefore, any code motion, such as hoisting, requires dependency analysis to ensure correctness.
+In contrast, the Sea of Nodes IR \cite{hotspot} does not explicitly impose program structure; instead, it captures all program structures implicitly through dependencies.
+This design allows for flexible code motion and local optimizations, with the program structure reconstructed during code generation.
+Inspired by \cite{rompf2010lightweight} and \cite{bravcevac2023graph}, Rhyme generates a loop-free, branch-free IR, capturing program structures via both data and control dependencies.
+While this design simplifies optimizations like loop hoisting and code scheduling, it complicates code generation, as the program structure must be reconstructed.
+The code generator needs to determine the order between loops, figure out how loops should be nested, and place instructions inside loops.
+Careless code placement, such as inserting an instruction inside a deeper loop nest, will cause redundant computations.
+This may also lead to incorrect results if the instruction corresponds to an aggregation in the source language.
+Therefore, the code generator must produce efficient code while respecting the semantics of the source language.\par
+
+In this paper, we introduce a novel code generation algorithm for Rhyme IR.
+It performs intelligent loop scheduling to derive an efficient loop structure for the program and utilizes a heuristic-guided algorithm to perform aggressive loop fusion.
+This optimization reduces extra loop control overhead, enables more optimization opportunities for the backend compiler, and reduces memory usage.
+The code generator guarantees program correctness by precomputing several program dependencies while respecting the semantic constraints of aggregations.
 \subsection{Paper Structure}
+\iffalse
 The paper is organized as follows: \par \par
+In section \ref{rhymeir}, we will look at the structure of Rhyme IR and its operators. This severs as a background for our code generation algorithm\par
+In section \ref{buildeps}, we will present several constraints that our code generator must follow to ensure program correctness.
+Then we will present the dependencies we need to pre-compute, these dependencies will be followed by the code generator during code scheduling to guarantee correctness.\par
+In section \ref{codeschedule}, we will dive deep into our code scheduling algorithm.
+We will investigate how the loop are scheduled and fused to improve performance.
+We will also look at the heuristic used to perform greedy scheduling.\par
+In section \ref{evaluation}, we compare the performance of old and new code generator of Rhyme, compare the generated code and explain the differences.\par
+In section \ref{conclusion}, we summarize the work and draw some conclusions.
+\fi
+The paper is organized as follows:\par \par
+In Section \ref{rhymeir}, we explore the structure of Rhyme IR and its operators, providing essential background for our code generation algorithm.\par
+In Section \ref{buildeps}, we present several constraints that our code generator must adhere to in order to ensure program correctness.
+We then detail the dependencies that need to be pre-computed, guiding the code generator during code scheduling to guarantee correctness.\par
+In Section \ref{codeschedule}, we delve into our code scheduling algorithm, examining how loops are scheduled and fused to improve performance.
+We also discuss the heuristic employed for greedy scheduling.\par
+In Section \ref{evaluation}, we compare the old and new code generators of Rhyme, compare the generated code and explain the differences.\par
+Finally, in Section \ref{conclusion}, we summarize our findings and draw conclusions.
+
+\section{Rhyme IR}\label{rhymeir}
+\iffalse
+Rhyme IR consists of two types of operators: \textit{generators} and \textit{assignments}:\begin{itemize}
+  \item Generators are operators responsible for iterating over data sources and intermediate results.
+Generator operations are transformed into loops in the generated code.
+User can use \inline{*} to represent a generator.
+For example, \inline{data.*} binds the generator \inline{*} to \inline{data}, it will iterate over \inline{data}, enumerating its values.
+  \item Assignments are operators that model all the computation for intermediate or final results.
+Rhyme generates temporaries in the IR to store intermediate results for sub-queries.
+For example, in query \inline{sum(data.*A.val) + sum(data.*B.val)}, \inline{sum(data.*A.val)} and \inline{sum(data.*B.val)} are stored in two temporaries - \inline{tmp[1]} and \inline{tmp[2]}.
+The final result is stored in \inline{tmp[0]}, and an assignment \inline{tmp[0] = tmp[1] + tmp[2]} is generated to compute the addition of the two sum.
+\end{itemize}
+As we mentioned above, intermediate results are stored in temporaries.
+Generators can iterate over temporaries, this will introduce a generator to temp dependencies.
+Also, since assignments can use temporaries as operand and write the result to temporaries, this create an assignment to temp dependencies.
+An assignment's right hand side (its operands) may also indexing into a data source/intermediate result by a generator.
+This gives us assignment to generator dependencies.\par
+Each assignment will write to a temporaries, i.e., its writeSym.
+There maybe multiple assignments write to a same temp.
+To ensure the correct ordering of those assignments and guarantee program correctness, Rhyme assigns write ranks to assignments that write to the same temp during IR generation.
+All assignments to the same temp must be ordered by their write rank.
+This gives us assignment to assignment dependencies.
+Since there maybe multiple assignments write to a temp, the temporary variable is not fully materialized until the last assignment that write to it.
+This introduces temp to assignment dependencies.\par
+Our brief discussions above already illustrate that the Rhyme IR models the complicated dependencies between generator, assignment and temp variables.
+When we generate code, we must faithful follow these dependencies in the IR to ensure program correctness.\par
+Besides the dependencies mentioned above, we also need to consider the semantic of aggregations in the IR.
+For temporaries that store aggregation results, we need to wait for it to be fully materialized in order to use it, otherwise we may use a partial aggregation and gives incorrect results.
+This will add extra dependencies in the IR level, we will talk more in detail about this in Section \ref{buildeps}.
+\fi
+Rhyme IR consists of two types of operators: \textit{generators} and \textit{assignments}:\begin{itemize}
+  \item Generators are operators responsible for iterating over data sources and intermediate results.
+Generator operations are transformed into loops in the generated code.
+Users can use \inline{*} to represent a generator.
+For example, \inline{data.*} binds the generator \inline{*} to \inline{data}, iterating over \inline{data} and enumerating its values.
+
+  \item Assignments are operators that model all computations for intermediate or final results.
+Rhyme generates temporaries in the IR to store intermediate results for sub-queries.
+For instance, in the query \inline{sum(data.*A.val) + sum(data.*B.val)}, \inline{sum(data.*A.val)} and \inline{sum(data.*B.val)} are stored in two temporaries - \inline{tmp[1]} and \inline{tmp[2]}.
+The final result is stored in \inline{tmp[0]}, and an assignment \inline{tmp[0] = tmp[1] + tmp[2]} is generated to compute the addition of the two sums.
+\end{itemize}
+As mentioned above, intermediate results are stored in temporaries.
+Generators can iterate over temporaries, introducing a generator-to-temp dependency.
+Additionally, since assignments can use temporaries as operands and write the result to temporaries, this creates an assignment-to-temp dependency.
+An assignment's right-hand side (its operands) may also index into a data source/intermediate result by a generator, resulting in assignment-to-generator dependencies.\par
+
+Each assignment will write to a temporary, i.e., its writeSym.
+There may be multiple assignments writing to the same temp.
+To ensure the correct ordering of those assignments and guarantee program correctness, Rhyme assigns write ranks to assignments that write to the same temp during IR generation.
+All assignments to the same temp must be ordered by their write rank.
+This results in assignment-to-assignment dependencies.
+Since there may be multiple assignments writing to a temp, the temporary variable is not fully materialized until the last assignment that writes to it.
+This introduces temp-to-assignment dependencies.\par
+
+Our brief discussions above illustrate that the Rhyme IR models the complex dependencies between generators, assignments, and temp variables.
+When generating code, we must faithfully follow these dependencies in the IR to ensure program correctness.\par
+
+In addition to the dependencies mentioned above, we also need to consider the semantics of aggregations in the IR.
+For temporaries that store aggregation results, we need to wait for them to be fully materialized before using them; otherwise, we may use a partial aggregation and obtain incorrect results.
+This introduces additional dependencies at the IR level, which we will discuss in more detail in Section \ref{buildeps}.
+\section{Semantic constraints and Dependency construction}\label{buildeps}
+\iffalse
+The code generator needs to follow several semantic constraints to ensure program correctness and generate efficient code.
+The code generator also needs to respect the dependencies between assignments and generators in th IR during code scheduling.
+In this section, we will delve into this two part.
+This will give us a deeper understanding of the intuition of our code scheduling algorithm.
+Since generators will be transformed into loops and assignments will be transformed into statements in the generated code.
+\par\textbf{Note: From now on, generator/loop and assignment/statement will be used interchangeably.}
+\subsection[]{Semantic constraints}
+We want to list several semantic constraints for the code generators here. These constraints will discipline the code generation for assignments and generators:
+\begin{enumerate}
+  \item \textbf{a statement will be only emitted once.} Since assignment model computations for intermediate and final results, they can be only emitted once.
+  \item \textbf{a loop maybe generated multiple times.} There could multiple assignments depend on one generator.
+The code generator can choose to either schedule them in one loop body or distribute them among multiple loops that correspond to the same generator.
+If a generator have n assignments depend on it.
+It can be generated at most n times.
+  \item \textbf{a statement will be emitted inside and only inside the loops it depends on.} This means we will never emit a statement inside irrelevant loops.
+Emitting an assignment inside irrelevant loops not only introduces extra overhead but may also gives incorrect results if the assignment computes an aggregation.
+  \item \textbf{statements and loops will be emitted while faithfully following the dependencies between them.} When emit statements and loops, the code generator must follow the dependencies between them to ensure program correctness.
+We will talk about this more in detail in subsection \ref{depsonstruct}.
+\end{enumerate}
+These four semantic constraints lay the foundation for our code generator.
+Any code generation algorithm following these constraints can guarantee the correctness of their generated program.
+A smart code generator, on the other hand, need to fuse different statements' loops together as much as possible to improve performance.
+\fi
+The code generator needs to follow several semantic constraints to ensure program correctness and generate efficient code.
+The code generator also needs to respect the dependencies between assignments and generators in the IR during code scheduling.
+In this section, we will delve into these two parts.
+This will give us a deeper understanding of the intuition behind our code scheduling algorithm.
+Since generators will be transformed into loops and assignments will be transformed into statements in the generated code.\par
+
+\textbf{Note: From now on, generator/loop and assignment/statement will be used interchangeably.}
+
+\subsection[]{Semantic constraints}\label{semconst}
+
+We want to list several semantic constraints for the code generator here. These constraints will discipline the code generation for assignments and generators:
+
+\begin{enumerate}
+\item \textbf{A statement will only be emitted once.} Since assignments model computations for intermediate and final results, they can only be emitted once.
+\item \textbf{A loop may be generated multiple times.} There could be multiple assignments depending on one generator.
+The code generator can choose to either schedule them in one loop body or distribute them among multiple loops that correspond to the same generator.
+If a generator has n assignments depending on it, it can be generated at most n times.
+\item \textbf{A statement will be emitted inside and only inside the loops it depends on.} This means we will never emit a statement inside irrelevant loops.
+Emitting an assignment inside irrelevant loops not only introduces extra overhead but may also give incorrect results if the assignment computes an aggregation.
+\item \textbf{Statements and loops will be emitted while faithfully following the dependencies between them.} When emitting statements and loops, the code generator must follow the dependencies between them to ensure program correctness.
+We will talk about this in more detail in subsection \ref{depsonstruct}.
+\end{enumerate}
+
+These four semantic constraints lay the foundation for our code generator.
+Any code generation algorithm following these constraints can guarantee the correctness of its generated program.
+However, these constraints do not enforce a strict program structure.
+Statements that are independent of each other can be scheduled in an arbitrary order.
+A generator can be emitted multiple times, so statements can be placed among them as long as the dependencies are followed.
+Therefore, a smart code generator needs to fuse different statements' loops together as much as possible to improve performance.
+\subsection[]{Dependency construction}\label{depsonstruct}
+\iffalse
+We briefly explored the dependencies of Rhyme IR in Section \ref{rhymeir}.
+We already observed a complicated dependency between assignments, generators and temporaries.
+Dependency to a temp can be inlined to the last assignment that update it.
+Therefore, we obtain four kinds of dependencies:
+\begin{enumerate}
+  \item \textbf{assignment-to-assignment} dependencies.
+These dependencies are introduced by the def-use effect of assignments.
+If assignment a uses temp1, and the last assignment that modifies temp1 is b, then a depends on b.
+Also, an assignment depends on the previous assignment that update the same temp (based on the write rank).
+  \item \textbf{assignment-to-generator} dependencies.
+If an assignment uses a generator (the rhs indexing into a temp by a generator or the lhs is grouped by the generator), the assignment depends on the generator.
+  \item \textbf{generator-to-assignment} dependencies. If a generator iterates over a temp, it depends on the last assignment that modifies this temp.
+  \item \textbf{generator-to-generator} dependencies.
+If the value that generator A iterates over is generated by another generator B, A depends on B.
+For example, in query \inline{data.*A.*B}, \inline{*A} depends on \inline{*B}.
+\end{enumerate}
+Dependencies are inherently transitive, if A depends on B and B depends on C, A also depends on C.
+According to our semantic constraints, assignment are only emitted once.
+Therefore, we use assignment/statement as the basic unit of dependencies.
+\par During dependencies construction phase, we first pre compute two types of dependencies:
+\begin{itemize}
+  \item \textbf{stmtdeps}: For each assignment/statement, this tracks all the assignments it transitively depends on, deriving from all four type of dependencies listed above.
+  \item \textbf{loopdeps}: For each assignment/statement, this tracks all the generators/loops it transitively depends on, i.e., the loop nests it resides in.
+When tracking loopdeps, we only follow assignment-to-generator and generator-to-generator dependencies.
+We never track loop dependencies beyond  assignments.
+The reason for it is simple: if assignment s1 depends on loop l1, and l1 depends on assignment s2.
+l1 only depends on the materialized result that s2 writes to.
+If s2 again depends on l2, the result is not fully materialized until l2 closes, therefore l1 does not need to be nested inside l2.
+As a result, s1 also does not depend on l2.
+\end{itemize}
+stmtdeps stores the dependencies between assignments, so we need to emit assignments following the topological order of this dependencies graph.
+For an assignment, when all prerequisites in stmtdeps are emitted, the assignment is ready to be emitted, although it may not have all loops opened at the moment.
+\par loopdeps then tells us the loops that an assignment need to be emitted in.
+An assignment can only be emitted when all loops it depends on are opened and all its prerequisites in stmtdeps are emitted.\par
+\fi
+We briefly explored the dependencies of Rhyme IR in Section \ref{rhymeir}.
+We already observed a complicated dependency between assignments, generators, and temporaries.
+Dependency to a temp can be inlined to the last assignment that updates it.
+Therefore, we obtain four kinds of dependencies:
+
+\begin{enumerate}
+\item \textbf{Assignment-to-assignment} dependencies: These dependencies are introduced by the def-use effect of assignments.
+If assignment $a$ uses $temp1$, and the last assignment that modifies $temp1$ is $b$, then $a$ depends on $b$.
+Also, an assignment depends on the previous assignment that updates the same temp (based on the write rank).
+\item \textbf{Assignment-to-generator} dependencies: If an assignment uses a generator (the rhs indexing into a temp by a generator or the lhs is grouped by the generator), the assignment depends on the generator.
+\item \textbf{Generator-to-assignment} dependencies: If a generator iterates over a temp, it depends on the last assignment that modifies this temp.
+\item \textbf{Generator-to-generator} dependencies: If the value that generator $A$ iterates over is generated by another generator $B$, $A$ depends on $B$.
+For example, in query \inline{data.*A.*B}, \inline{*B} depends on \inline{*A}.
+\end{enumerate}
+
+Dependencies are inherently transitive: if $A$ depends on $B$ and $B$ depends on $C$, $A$ also depends on $C$.
+According to our semantic constraints, assignments are only emitted once.
+Therefore, we use assignment/statement as the basic unit of dependencies.\par
+
+During the dependencies construction phase, we first pre-compute two types of dependencies:
+
+\begin{itemize}
+\item \textbf{Stmtdeps}: For each assignment/statement, this tracks all the assignments it transitively depends on, derived from all four types of dependencies listed above.
+\item \textbf{Loopdeps}: For each assignment/statement, this tracks all the generators/loops it transitively depends on, i.e., the loop nests it resides in.
+When tracking loopdeps, we only follow assignment-to-generator and generator-to-generator dependencies.
+We never track loop dependencies beyond assignments.
+The reason for this is simple: if assignment $s1$ depends on loop $l1$, and $l1$ depends on assignment $s2$, $l1$ only depends on the materialized result that $s2$ writes to.
+If $s2$ again depends on $l2$, the result is not fully materialized until $l2$ closes, therefore $l1$ does not need to be nested inside $l2$.
+As a result, $s1$ also does not depend on $l2$.
+\end{itemize}
+
+Stmtdeps stores the dependencies between assignments, so we need to emit assignments following the topological order of this dependency graph.
+For an assignment, when all prerequisites of it in stmtdeps are emitted, the assignment is ready to be emitted, although it may not have all loops opened at the moment.\par
+
+Loopdeps then tell us the loops that an assignment needs to be emitted in.
+An assignment can only be emitted when all loops it depends on are opened and all its prerequisites in stmtdeps are emitted.\par
+
+\iffalse
+Although stmtdeps and loopdeps already give us a lot of information about the program, it still does not fully model all the required dependencies.
+Rhyme is a query language by design, therefore it supports different types of aggregations - sum, count, min, max, etc.
+If an assignment computes the result of an aggregation and stores the result in a temp.
+The aggregation is only fully materialized until all the loops it depends on are all closed, i.e., the aggregation is fully computed.
+Any assignment that uses this temp will need to wait for the aggregation to be fully materialized, otherwise it may uses a partial aggregation and produce incorrect results.\par
+This observation inspires two extra dependencies we need to track to ensure program correctness:
+\begin{itemize}
+  \item \textbf{stmt2stmtByLoop}: if statement s1 depends on s2, this tracks the loops of s2 that need to be closed for s1 to be emittable.
+This is needed for aggregation statements.
+We need to wait for all dependent loops to close for the aggregation to be fully materialized.
+  \item \textbf{stmt2stmtloopAfterloop}: if assignment s1 depends on s2, this tracks the pair of loops (l1, l2) such that: s1 depends on l1, s2 depends on l2 and s1's l1 needed to be emitted after s2's l2.
+This gives an ordering of loops and will be used in the heuristic for loop scheduling.
+\end{itemize}
+We currently only track stmt2stmtByLoop and stmt2stmtloopAfterloop between statements that write to different temporary variables.
+If s1 depends on s2, we compute these two dependencies as follows:\par
+If s2 depends on loop l and s1 does not.
+Then s1 need to wait for s2's l loop to close so that the temp s2 writes to is fully materialized.
+So we need to set \inline{stmt2stmtByLoop[s1][s2][l] = true}.
+If s1 depends on l, s1 may or may not need to wait for s2's l to close, depending on the computation s2 models and how s1 uses s2's result.
+For example, if s2 writes to \inline{temp1[*l]} and s1 uses \inline{temp1[*l]}, s1 and s2 can be put in the same loop body for l.
+However, we can perform arithmetic operations on the generators in Rhyme.
+Since generator represents iterator, we can offset the generator for indexing.
+So, if s2 writes to \inline{temp1[*l]} and s1 uses \inline{temp1[func(*l)]}, s1 may need to wait for \inline{temp1} to be fully materialized to use it.
+Therefore, s1 and s2 may or may not be able to be put in the same l loop, depending on the semantic of \inline{func}.
+With more precise dependency analysis and reasoning, we can relax the requirements for stmt2stmtByLoop and explore more optimization opportunities.
+stmt2stmtByLoop offers a precise, fine-grained dependencies between statements with regard to loops.
+It gives the compiler more information about how loops should be scheduled.
+And the compiler construct stmt2stmtByLoop to reflect the semantic of different assignments and the dependencies between them.
+\fi
+Although \texttt{stmtdeps} and \texttt{loopdeps} already provide a lot of information about the program, they still do not fully model all the required dependencies.
+Rhyme is a query language by design; therefore, it supports different types of aggregations - sum, count, min, max, etc.
+If an assignment computes the result of an aggregation and stores the result in a temporary variable, the aggregation is only fully materialized when all the loops it depends on are closed, i.e., the aggregation is fully computed.
+Any assignment that uses this temporary variable will need to wait for the aggregation to be fully materialized; otherwise, it may use a partial aggregation and produce incorrect results.\par
+This observation inspires two extra dependencies we need to track to ensure program correctness:
+\begin{itemize}
+\item \textbf{stmt2stmtByLoop}: if statement \texttt{s1} depends on \texttt{s2}, this tracks the loops of \texttt{s2} that need to be closed for \texttt{s1} to be emittable.
+This is needed for aggregation statements.
+We need to wait for all dependent loops to close for the aggregation to be fully materialized.
+\item \textbf{stmt2stmtloopAfterloop}: if assignment \texttt{s1} depends on \texttt{s2}, this tracks the pair of loops (\texttt{l1}, \texttt{l2}) such that: \texttt{s1} depends on \texttt{l1}, \texttt{s2} depends on \texttt{l2} and \texttt{s1}'s \texttt{l1} needs to be emitted after \texttt{s2}'s \texttt{l2}.
+This gives an ordering of loops and will be used in the heuristic for loop scheduling.
+\end{itemize}
+We currently only track \texttt{stmt2stmtByLoop} and \texttt{stmt2stmtloopAfterloop} between statements that write to different temporary variables.
+If \texttt{s1} depends on \texttt{s2}, we compute these two dependencies as follows:\par
+If \texttt{s2} depends on loop \texttt{l} and \texttt{s1} does not, then \texttt{s1} needs to wait for \texttt{s2}'s \texttt{l} loop to close so that the temporary variable \texttt{s2} writes to is fully materialized.
+So we need to set \inline{stmt2stmtByLoop[s1][s2][l] = true}.
+If \texttt{s1} depends on \texttt{l}, \texttt{s1} may or may not need to wait for \texttt{s2}'s \texttt{l} to close, depending on the computation \texttt{s2} models and how \texttt{s1} uses \texttt{s2}'s result.
+For example, if \texttt{s2} writes to \texttt{temp1[*l]} and \texttt{s1} uses \texttt{temp1[*l]}, \texttt{s1} and \texttt{s2} can be put in the same loop body for \texttt{l}.
+However, we can perform arithmetic operations on the generators in Rhyme.
+Since a generator represents an iterator, we can offset the generator for indexing.
+So, if \texttt{s2} writes to \texttt{temp1[*l]} and \texttt{s1} uses \texttt{temp1[func(*l)]}, \texttt{s1} may need to wait for \texttt{temp1} to be fully materialized to use it.
+Therefore, \texttt{s1} and \texttt{s2} may or may not be able to be put in the same \texttt{l} loop, depending on the semantics of \texttt{func}.
+With more precise dependency analysis and reasoning, we can relax the requirements for \texttt{stmt2stmtByLoop} and explore more optimization opportunities.
+\texttt{stmt2stmtByLoop} offers precise, fine-grained dependencies between statements with regard to loops.
+It gives the compiler more information about how loops should be scheduled.
+And the compiler constructs \texttt{stmt2stmtByLoop} to reflect the semantics of different assignments and the dependencies between them.\par
+\iffalse
+stmt2stmtloopAfterloop, on the other hand, is derived from stmt2stmtByLoop.
+And it is constructed to hint an ordering of loops.
+If \inline{stmt2stmtByLoop[s1][s2][l2] = true}, we know that s1 need to wait for s2'l2 loop to close.
+This means s1 needs to be emitted after s2's l2 loop closes.
+If s1 depends on l1 and s2 does not depend on l1.
+s1's l1 loop needs to be scheduled after s2's l2 loop, and we set \inline{stmt2stmtloopAfterloop[s1][s2][l1][l2] = true}.
+We never emit a statement inside irrelevant loops according to the semantic constraints described in subsection \ref{semconst}.Since s1 does not depend on l2 and s2 does not depend on l1.
+We will never emit s1 inside l2 and emit s2 inside l1, s1's l1 and s2's l2 can not be nested within each other.
+Since we need to emit s1 after s2's l2, s1'l1 must also be scheduled after s2's l2.
+This enforecs an ordering between l1 and l2 per s1 and s2.
+Whenever we have a list of statements that are ready to be issued, we can merge all the per statement pair loop order among these statements into a local loopAfterLoop relation.
+This local loopAfterLoop relation tells us the order by which we should emit loops.
+This will be used as a hint for loop scheduling, which we will talk more in detail in Section \ref{codeschedule}.
+\fi
+\texttt{stmt2stmtloopAfterloop}, on the other hand, is derived from \texttt{stmt2stmtByLoop}, and it is constructed to hint an ordering of loops.
+If \inline{stmt2stmtByLoop[s1][s2][l2] = true}, we know that \texttt{s1} needs to wait for \texttt{s2}'s \texttt{l2} loop to close.
+This means \texttt{s1} needs to be emitted after \texttt{s2}'s \texttt{l2} loop closes.
+If \texttt{s1} depends on \texttt{l1} and \texttt{s2} does not depend on \texttt{l1}, \texttt{s1}'s \texttt{l1} loop needs to be scheduled after \texttt{s2}'s \texttt{l2} loop, and we set \inline{stmt2stmtloopAfterloop[s1][s2][l1][l2] = true}.
+We never emit a statement inside irrelevant loops according to the semantic constraints described in subsection \ref{semconst}.
+Since \texttt{s1} does not depend on \texttt{l2} and \texttt{s2} does not depend on \texttt{l1}, we will never emit \texttt{s1} inside \texttt{l2} and emit \texttt{s2} inside \texttt{l1}.
+\texttt{s1}'s \texttt{l1} and \texttt{s2}'s \texttt{l2} can not be nested within each other.
+Since we need to emit \texttt{s1} after \texttt{s2}'s \texttt{l2}, \texttt{s1}'s \texttt{l1} must also be scheduled after \texttt{s2}'s \texttt{l2}.
+This enforces an ordering between \texttt{l1} and \texttt{l2} per \texttt{s1} and \texttt{s2}.
+Whenever we have a list of statements that are ready to be issued, we can merge all the per statement-pair loop order among these statements into a local \texttt{loopAfterLoop} relation.
+This local \texttt{loopAfterLoop} relation tells us the order by which we should emit loops.
+This will be used as a hint for loop scheduling, which we will talk about in more detail in Section \ref{codeschedule}.\par
+In this section, we delve deep into the semantic constraints and dependency construction of our code generator.
+The constraints and dependencies form the basis of our code generation algorithm to ensure correctness.
+We will discuss the loop scheduling algorithm in more detail in the next section.
+\section{Code scheduling}\label{codeschedule}
+\section{Evaluation}\label{evaluation}
+\section{Conclusion}\label{conclusion}
 %\newpage

### 2024-05-05 paper.tex
finish evaluation

@@
 \usepackage{mathpartir}
+\usepackage{float}
 \input{macros}
@@
 \section{Code scheduling}\label{codeschedule}
+\iffalse
+Semantic constraints and pre-computed dependencies discipline code generation, enforcing restrictions on the code generator to guarantee correctness.
+They decide which statements should be scheduled before a statement.
+They restrict which loops a statement need to reside in.
+They also decide what loops need to be closed for a statement to be issuable.
+However, they do not restrict a program structure.
+And they leave the code generator a lot of freedom to schedule statements and loops.
+\par
+As we mentioned in subsection \ref{semconst}, a loop can be emitted multiple times.
+Although we want to put all statements that depend on a generator in one loop body, sometimes it is infeasible due to the aforementioned \texttt{stmt2stmtByLoop} and \texttt{stmt2stmtloopAfterloop} dependencies.
+For example, consider the following IR snippet:
+\begin{lstlisting}[numbers=left]
+  tmp[1] += data1[*A]['val1']
+  tmp[2] += (data2[*B]['val2'] / tmp[1])
+  tmp[3][*A] = data1[*A]['val3'] / tmp[2]
+  \end{lstlisting}
+The semantic meaning of this snippet is not irrelevant here, we only care about the dependencies captured in the IR.
+Assignment 1 uses *A in the rhs, so it depends on *A, same thing for assignment 3, and assignment 2 depends on *B.
+Assignment 3 depends on assignment 2, and assignment 2 depends on assignment 1 due to the def-use chain.
+Since both assignment 1 and 2 compute an aggregation (sum), their loop (*A and *B) need to be closed for tmp[1] and tmp[2] to be fully materialized.
+Therefore assignment 2 need to wait for assignment 1's *A loop to close, assignment 3 need to wait for assignment 2's *B loop to close.\par
+Ideally, we want to put both assignment 1 and 3 in the same *A loop to reduce the loop control overhead, i.e., fuse assignment 1 and 3's *A loop.
+However, this is not possible due to the extra dependency of assignment 2.
+Assignment 2 needs to be emitted after assignment 1.
+So assignment 2's *B loop should be emitted after assignment 1's *A loop closes.
+Similarly, assignment 3's *A loop need to be after assignment 2's *B loop.
+So assignment 3's *A loop can only be scheduled after assignment1's *A loop closes.
+This prevents assignment 1 and 3 from being placed in the same *A loop.\par
+For large and complex applications, there are multiple fusion opportunities when scheduling code.
+And the compiler must carefully schedule loops to maximize fusion.
+For example, consider the following IR snippet:
+\begin{lstlisting}[numbers=left]
+  tmp[1][*A] += data1[*A][*B]['val1']
+  tmp[2][*A] = tmp[1][*A] + data2[*A]['val2']
+  tmp[3][*A] = tmp[1][*A] + data3[*A]['val3']
+  \end{lstlisting}
+Assignment 1 resides in both *A and *B loops, assignment 2 and 3 only resides in *A loop.
+Also, assignment 2 and 3 need to be scheduled after assignment 1 due to dependencies.
+We we emit assignment 1, if we choose to first emit *B loop and then emit *A loop inside *B loop.
+We can not put the succeeding assignments inside the nested *A loop as these assignments are only dependent on the *A loop, so they can not be nested within *B.
+Therefore, the compiler needs to close the *B and *A loop nest.
+Then reopen *A loop to emit assignment 2 and 3.
+If we choose to emit *A loop first when emitting assignment 1, this redundance will be eliminated and we can fuse assignment 1,2,3's *A loop.\par
+As we observed before, the code generator need to perform non-trivial reasoning to generate optimal loop structure for the IR while satisfy the dependencies.
+The goal is to maximize loop fusion to improve performance.
+We now present our heuristic-guided loop scheduling algorithm that can perform aggressive loop fusion efficiently.\par
+\fi
+Semantic constraints and pre-computed dependencies discipline code generation, enforcing restrictions on the code generator to guarantee correctness.
+They determine which statements should be scheduled before others, restrict the loops in which a statement can reside, and dictate which loops need to be closed for a statement to be issuable.
+However, they do not restrict the program structure, leaving the code generator with considerable freedom to schedule statements and loops.
+
+As mentioned in subsection \ref{semconst}, a loop can be emitted multiple times.
+Although we aim to place all statements that depend on a generator in one loop body, sometimes it is infeasible due to the aforementioned \texttt{stmt2stmtByLoop} and \texttt{stmt2stmtloopAfterloop} dependencies.
+For example, consider the following IR snippet:
+\begin{lstlisting}[numbers=left]
+tmp[1] += data1[*A]['val1']
+tmp[2] += (data2[*B]['val2'] / tmp[1])
+tmp[3][*A] = data1[*A]['val3'] / tmp[2]
+\end{lstlisting}
+The semantic meaning of this snippet is not relevant here; we only care about the dependencies captured in the IR.
+Assignment 1 uses \texttt{*A} in the right-hand side (RHS), so it depends on \texttt{*A}; the same goes for assignment 3, and assignment 2 depends on \texttt{*B}.
+Assignment 3 depends on assignment 2, and assignment 2 depends on assignment 1 due to the def-use chain.
+Since both assignment 1 and 2 compute an aggregation (sum), their loops (\texttt{*A} and \texttt{*B}) need to be closed for \texttt{tmp[1]} and \texttt{tmp[2]} to be fully materialized.
+Therefore, assignment 2 needs to wait for assignment 1's \texttt{*A} loop to close, and assignment 3 needs to wait for assignment 2's \texttt{*B} loop to close.
+
+Ideally, we want to put both assignment 1 and 3 in the same \texttt{*A} loop to reduce the loop control overhead, i.e., fuse assignment 1 and 3's \texttt{*A} loop.
+However, this is not possible due to the extra dependency of assignment 2.
+Assignment 2 needs to be emitted after assignment 1.
+So assignment 2's \texttt{*B} loop should be emitted after assignment 1's \texttt{*A} loop closes.
+Similarly, assignment 3's \texttt{*A} loop needs to be after assignment 2's \texttt{*B} loop.
+Thus, assignment 3's \texttt{*A} loop can only be scheduled after assignment 1's \texttt{*A} loop closes, preventing assignment 1 and 3 from being placed in the same \texttt{*A} loop.
+
+For large and complex applications, there are multiple fusion opportunities when scheduling code.
+The compiler must carefully schedule loops to maximize fusion.
+For example, consider the following IR snippet:\par
+\begin{figure}[H]
+\begin{lstlisting}[numbers=left]
+tmp[1][*A] += (data0[*A]['val0'] + data1[*B]['val1'])
+tmp[2][*A] = tmp[1][*A] + data2[*A]['val2']
+tmp[3][*A] = tmp[1][*A] + data3[*A]['val3']
+\end{lstlisting}
+\caption{}
+\label{loopheu}
+\end{figure}
+Assignment 1 resides in both \texttt{*A} and \texttt{*B} loops, while assignment 2 and 3 only reside in the \texttt{*A} loop.
+Also, assignment 2 and 3 need to be scheduled after assignment 1 due to dependencies.
+When we emit assignment 1, if we first emit the \texttt{*B} loop and then the \texttt{*A} loop inside the \texttt{*B} loop, we cannot put the succeeding assignments inside the nested \texttt{*A} loop as these assignments are only dependent on the \texttt{*A} loop, so they cannot be nested within \texttt{*B}.
+Therefore, the compiler needs to close the \texttt{*B} and \texttt{*A} loop nest and then reopen the \texttt{*A} loop to emit assignment 2 and 3.
+If we choose to emit the \texttt{*A} loop first when emitting assignment 1, this redundancy will be eliminated, and we can fuse assignment 1, 2, and 3's \texttt{*A} loop.
+
+As observed before, the code generator needs to perform non-trivial reasoning to generate an optimal loop structure for the IR while satisfying the dependencies.
+The goal is to maximize loop fusion to improve performance.
+We now present our heuristic-guided loop scheduling algorithm, which can perform aggressive loop fusion efficiently.
+\subsection*{Loop scheduling algorithm}
+\iffalse
+We use a heuristic guided approach to perform efficient loop scheduling. As we mentioned before, we use statement as the basic unit for code scheduling. \par
+\paragraph*{Iteratively emitting statements} If all prerequisite statements of a statement are emitted and have their loops (with respect to stmt2stmtByLoop) closed, we call this statement \textbf{ready}.
+If all dependent loops of a statement are opened, we call this statement \textbf{inloops}.
+If a statement is both ready and inloops, we call this statement \textbf{emittable}, and we can emit the statement.
+Since we have pre computed both stmtdeps, loopdeps and stmt2stmtByLoop, the first stage is trivial.
+We check for all remaining statements whether they are emittable, and emit them if so.
+We will iteratively emit statements until there are no emittable statements left.\par
+Although there are no emittable statements, some statements may still be ready, they just do not have all their dependent loops opened.
+If there are no ready statements, this means some statement are waiting for its prerequisites' loops to close.
+Then we close the current loop and return to the parent loop scope.\par
+ If there are several ready statements and they depend on multiple loops, we need to choose which loop to open next carefully.
+Careless choice of which loop to open next may eliminate future opportunities for loop fusion and generate suboptimal code.
+We also may take stmt2stmtloopAfterloop in to consideration, and avoid closing a loop halfway due to stmt2stmtByLoop, and reopen it later.
+To pick the optimal loop to open next, we use a greedy heuristic.\par
+\subsection*{Heuristic for opening new loop}
+We use a greedy heuristic to choose which loop to open next.
+When we have a set of ready statements, we first compute the set of loops that are currently unopened and are dependent by ready statements.
+Opening any of these loops will enable the opportunity for some ready statement to become emittable.
+Next, we compute a local loopAfterLoop relation by merging all the per-pair stmt2stmtloopAfterloop among un-emitted statements.
+This relation hints a loop order we need to follow when opening new loops.
+We do not want to open a loop and close it halfway due to the stmt2stmtByLoop dependencies.
+Then, we pick the loop that do not need to be scheduled after any other loops.
+If there is only one such loop, we just pick this loop and open it.
+\par
+If there are multiple such loops, we need to use a greedy heuristic to pick the best candidate.
+If there are multiple loops that can be opened, we want to open one that can host the most number of statements.
+If those loops are nested, we want to put the best candidate in the outer level, avoiding closing a loop halfway due to semantic constraints as discussed in Figure \ref{loopheu}.\par
+Our heuristic uses two statistics to pick the best candidate: \texttt{stmtCnt} and \texttt{readyStmtCnt}.
+For each loop candidate, stmtCnt computes the number of un-emitted statements that depend on it.
+readyStmtCnt computes the number of ready statements that depend on it.
+The heuristic first picks the loop candidate with a larger readyStmtCnt.
+In time of a tie, the heuristic picks the loop candidate with a larger stmtCnt.\par
+\paragraph*{Recurse into inner loop scope} After the heuristic pick the next loop to open, we open this new loop, and recurse inside the new loop scope.
+As the semantic constraints require all statements to be emitted only inside their dependent loops, we filter out the statements that do not depend on the new loop when we recurse.
+This can avoid place statement inside irrelevant loops.\par
+After we return from the recursion, we can repeat all stages again as some statements will become emittable after the recursion.
+We will again emit all emittable statements, then open a new loop to recurse.
+We will keep iterating until all statements are emitted.
+\fi
+We utilize a heuristic-guided approach to perform efficient loop scheduling. As mentioned earlier, we use statements as the basic unit for code scheduling.
+
+\paragraph*{Iteratively Emitting Statements} If all prerequisite statements of a statement are emitted and have their loops (with respect to \texttt{stmt2stmtByLoop}) closed, we call this statement \textbf{ready}.
+If all dependent loops of a statement are opened, we call this statement \textbf{inloops}.
+If a statement is both ready and inloops, we call this statement \textbf{emittable}, and we can emit the statement.
+Since we have pre-computed both \texttt{stmtdeps}, \texttt{loopdeps}, and \texttt{stmt2stmtByLoop}, the first stage is trivial.
+We check for all remaining statements whether they are emittable and emit them if so.
+We will iteratively emit statements until there are no emittable statements left.\par
+
+Although there are no emittable statements, some statements may still be ready; they just do not have all their dependent loops opened.
+If there are no ready statements, it means some statements are waiting for their prerequisites' loops to close.
+Then we close the current loop and return to the parent loop scope.\par
+
+If there are several ready statements and they depend on multiple loops, we need to carefully choose which loop to open next.
+Careless choice of which loop to open next may eliminate future opportunities for loop fusion and generate suboptimal code.
+We also take \texttt{stmt2stmtloopAfterloop} into consideration and avoid closing a loop halfway due to \texttt{stmt2stmtByLoop} and reopening it later.
+To pick the optimal loop to open next, we use a greedy heuristic.\par
+
+\subsection*{Heuristic for Opening New Loop}
+We employ a greedy heuristic to choose which loop to open next.
+When we have a set of ready statements, we first compute the set of loops that are currently unopened and are dependent on by ready statements.
+Opening any of these loops will enable the opportunity for some ready statement to become emittable.
+Next, we compute a local \texttt{loopAfterLoop} relation by merging all the per-pair \texttt{stmt2stmtloopAfterloop} among un-emitted statements.
+This relation indicates a loop order we need to follow when opening new loops.
+We do not want to open a loop and close it halfway due to the \texttt{stmt2stmtByLoop} dependencies.
+Then, we pick the loop that does not need to be scheduled after any other loops.
+If there is only one such loop, we pick this loop and open it.\par
+
+If there are multiple such loops, we need to use a greedy heuristic to pick the best candidate.
+If multiple loops can be opened, we want to open one that can host the most number of statements.
+If those loops are nested, we want to put the best candidate in the outer level, avoiding closing a loop halfway due to semantic constraints as discussed in Figure \ref{loopheu}.\par
+
+Our heuristic uses two statistics to pick the best candidate: \texttt{stmtCnt} and \texttt{readyStmtCnt}.
+For each loop candidate, \texttt{stmtCnt} computes the number of un-emitted statements that depend on it.
+\texttt{readyStmtCnt} computes the number of ready statements that depend on it.
+The heuristic first picks the loop candidate with a larger \texttt{readyStmtCnt}.
+In the case of a tie, the heuristic picks the loop candidate with a larger \texttt{stmtCnt}.\par
+
+\paragraph*{Recurse into Inner Loop Scope} After the heuristic picks the next loop to open, we open this new loop and recurse inside the new loop scope.
+As the semantic constraints require all statements to be emitted only inside their dependent loops, we filter out the statements that do not depend on the new loop when we recurse.
+This can avoid placing statements inside irrelevant loops.\par
+
+After we return from the recursion, we can repeat all stages again as some statements will become emittable after the recursion.
+We will again emit all emittable statements, then open a new loop to recurse.
+We will keep iterating until all statements are emitted.\par
+
+Our heuristic-guided loop scheduling algorithm greedily opens new loops to maximize loop fusion at code generation time.
+The code generator fuses the loops of different statements to improve performance.
+It generates efficient JavaScript code for loop-free, branch-free Rhyme IR, and then uses the backend compiler to compile the generated code into an executable format.
 \section{Evaluation}\label{evaluation}
+\iffalse
+In this section, we compare the old code generator and our new code generator. The old and new code generator differs in the following aspects:
+\begin{enumerate}
+  \item \textbf{Dependency granularity}: the old code generator only tracks dependencies at the temporary variable level while our new code generator track dependencies at the assignment level.
+This means our new code generator models the dependency more precisely and therefore have enabled more opportunities for code motion and scheduling.
+For example, in the old code generator, if both s1 and s2 writes to tmp1, and s1 needs to wait for s3's l loop to close to be emittable.
+s2 also needs wait for s3's l loop to close, regardless of the computation s2 represents (even if s2 is an unconditional initialization).\par
+  Also, the old code generator computes a global loopAfterLoop relation, while the new code generator computes a per statement-pair loopAfterLoop relation.
+The new loop scheduling algorithm computes a local loopAfterLoop relation based on the currently in-scope statements.
+Therefore, this loop ordering is more precise and up-to-date.
+This removes the outdated loop order restrictions in the old code generator and gives the compiler more flexibility to schedule loops.
+  \item \textbf{Semantic constraint}: the old code generator only allows the each generator to be emitted once.
+When the compiler could not make progress, It will emit the statements one by one without performing any loop fusion.
+In contrast, the new code generator by design allows a generator to be emitted multiple times as long as the semantic of the program is respected.
+The code generator will try to maximize fusing different statements' loops aggressively to improve performance.\par
+  The old generator also allow statements to be emitted inside irrelevant loops, which may sometimes gives incorrect results.
+\end{enumerate}
+\fi
+In this section, we compare the old code generator with our new code generator. The old and new code generators differ in the following aspects:
+
+\begin{enumerate}
+\item \textbf{Dependency granularity}: The old code generator only tracks dependencies at the temporary variable level, while our new code generator tracks dependencies at the assignment level.
+This means our new code generator models dependencies more precisely, enabling more opportunities for code motion and scheduling.
+For example, in the old code generator, if both $s1$ and $s2$ write to $tmp1$, and $s1$ needs to wait for $s3$'s loop to close to be emittable, $s2$ also needs to wait for $s3$'s loop to close, regardless of the computation $s2$ represents (even if $s2$ is an unconditional initialization of $tmp1$).
+
+Additionally, the old code generator computes a global \texttt{loopAfterLoop} relation, while the new code generator computes a per-statement-pair \texttt{loopAfterLoop} relation.
+The new loop scheduling algorithm computes a local \texttt{loopAfterLoop} relation based on the currently in-scope statements.
+Therefore, this loop ordering is more precise and up-to-date.
+This removes the outdated loop order restrictions in the old code generator and gives the compiler more flexibility to schedule loops.
+
+\item \textbf{Semantic constraint}: The old code generator only allows each generator to be emitted once.
+When the compiler could not make progress, it would emit the statements one by one without performing any loop fusion.
+In contrast, the new code generator, by design, allows a generator to be emitted multiple times as long as the semantics of the program are preserved.
+The code generator will try to maximize fusing different statements' loops aggressively to improve performance.
+
+The old generator also allows statements to be emitted inside irrelevant loops, which may sometimes yield incorrect results.
+\end{enumerate}
+
 \section{Conclusion}\label{conclusion}

### 2024-05-06 paper.tex
finish draft

@@
 \end{enumerate}
+\iffalse
+We compare the generated code of both the old and the new code generator on the Rhyme solution for advent of code day5-part1.
+The old code generator already generates decent code, and the new code generator generates similar code to the old code generator in most test cases.
+The old generator only fails at certain complex applications.
+Therefore, we choose advent of code day5-part1 for comparison as the old code generator fails to schedule all statements for this application and uses repeated generators to generate the remaining IR.\par
+The JavaScript code that the old code generator generates is 223 lines while the new code generator only generates 134 lines.
+This is $39.9\%$ reduce in the code size.
+The new code generator manages to merge more statements' loops together, therefore reduces the code size.
+When executing on the puzzle inputs from advent of code, the execution time reduces by $21.6\%$.\par
+In conclusion, compared with the old code generator, our new code generator generates more efficient code by aggressively performing loop fusion at code generation time.
+\fi
+We compare the generated code of both the old and the new code generator on the Rhyme solution for Advent of Code day5-part1.
+The old code generator already generates decent code, and the new code generator generates similar code to the old code generator in most test cases.
+The old generator only fails at certain complex applications.
+Therefore, we choose Advent of Code day5-part1 for comparison as the old code generator fails to schedule all statements for this application and uses repeated generators to generate the remaining IR.\par
+
+The JavaScript code that the old code generator generates is 223 lines while the new code generator only generates 134 lines.
+This is a $39.9\%$ reduction in the code size.
+The new code generator manages to merge more statements' loops together, therefore reducing the code size.
+When executing on the puzzle inputs from Advent of Code, the execution time reduces by $21.6\%$.\par
+
+In conclusion, compared with the old code generator, our new code generator generates more efficient code by aggressively performing loop fusion at code generation time.
 
 \section{Conclusion}\label{conclusion}
-%\newpage
+\iffalse
+In this paper, we present an efficient code generation algorithm for loop-free, branch-free Rhyme IR.
+We present the constraints and dependencies that the code generator must follow in order to guarantee the correctness of generated code for Rhyme IR.
+We also present a heuristic guided loop scheduling algorithm that can achieve loop fusion at code generation time.
+\fi
+In this paper, we present an efficient code generation algorithm for loop-free, branch-free Rhyme IR.
+We present the constraints and dependencies that the code generator must follow to guarantee the correctness of the generated code for Rhyme IR.
+We also present a heuristic-guided loop scheduling algorithm that can achieve loop fusion at code generation time.
+\newpage
 \bibliographystyle{acm}

### 2024-05-06 paper.tex
add figure

@@
 Therefore, we choose Advent of Code day5-part1 for comparison as the old code generator fails to schedule all statements for this application and uses repeated generators to generate the remaining IR.\par
-
-The JavaScript code that the old code generator generates is 223 lines while the new code generator only generates 134 lines.
+\begin{figure}[H]
+  \centering
+      \includegraphics[scale=0.57]{evaluation.pdf}
+      \caption{Old vs New code generator}
+      \label{evaluationpic}
+  \end{figure}
+As shown in Figure \ref{evaluationpic}, the JavaScript code that the old code generator generates is 223 lines while the new code generator only generates 134 lines.
 This is a $39.9\%$ reduction in the code size.
@@
 We also present a heuristic-guided loop scheduling algorithm that can achieve loop fusion at code generation time.
-\newpage
+%\newpage
 \bibliographystyle{acm}
